Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-27 background.tex
work on related work, justifying modalities. not done with either, but progress.

In typical system configurations, all memory addresses seen by programs running on modern computers are \emph{virtualized}: the address observed by a running
program generally will not correspond directly to the physical location in memory, and may not even correspond to a physical location that \emph{exists} 
in the machine. Instead, these \emph{virtual} addresses are translated to \emph{physical} addresses that correspond directly to locations in RAM. On most 
modern architectures, this translation is performed through cooperation of the hardware and OS kernel: while executing an instruction that dereferences a 
(virtual) address, the CPU's \emph{memory management unit} (\textsc{MMU}) hardware performs \emph{address translation},
resulting in a physical address used to access the cache\footnote{Technically, for performance reasons most caches are indexed with parts of the virtual 
address, but tagged with the physical data addresses, so cache lookups and address translations can proceed in parallel.} and/or memory-bus.
\setlength{\columnseprule}{4pt}
\begin{figure}\footnotesize
%    \begin{framed}
%\column{.59\textwidth}    
\begin{subfigure}\footnotesize

@@

\begin{tikzpicture}[x=0.75pt,y=0.75pt,yscale=-0.75,xscale=0.75]
%uncomment if require: \path (0,427); %set diagram left start at 0, and has height of 427

@@

\draw   (33,141) -- (123,141) -- (123,292) -- (33,292) -- cycle ;

@@

\draw   (166,141) -- (256,141) -- (256,292) -- (166,292) -- cycle ;

@@

\draw   (295,141) -- (385,141) -- (385,292) -- (295,292) -- cycle ;

@@

\draw   (427,141) -- (517,141) -- (517,292) -- (427,292) -- cycle ;

@@

\draw   (115,54) -- (185,54) -- (185,94) -- (115,94) -- cycle ;

@@

\draw   (185,54) -- (255,54) -- (255,94) -- (185,94) -- cycle ;

@@

\draw   (255,54) -- (325,54) -- (325,94) -- (255,94) -- cycle ;

@@

\draw   (325,54) -- (395,54) -- (395,94) -- (325,94) -- cycle ;

@@

\draw   (395,54) -- (465,54) -- (465,94) -- (395,94) -- cycle ;

@@

\draw   (465,54) -- (535,54) -- (535,94) -- (465,94) -- cycle ;

@@

\draw    (211.5,96) .. controls (-36.44,98.94) and (15.75,190.24) .. (30.17,213.65) ;
\draw [shift={(31,215)}, rotate = 238.24] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw   (33,191) -- (123,191) -- (123,231) -- (33,231) -- cycle ;

@@

\draw    (289.5,95) .. controls (115.15,110.76) and (137.77,175.03) .. (163.33,197.98) ;
\draw [shift={(164.5,199)}, rotate = 220.24] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw   (166,177) -- (256,177) -- (256,217) -- (166,217) -- cycle ;

@@

\draw  [dash pattern={on 4.5pt off 4.5pt}] (295,156) -- (385,156) -- (385,196) -- (295,196) -- cycle ;

@@

\draw   (427,212) -- (517,212) -- (517,252) -- (427,252) -- cycle ;

@@

\draw [color={rgb, 255:red, 0; green, 0; blue, 0 }  ,draw opacity=1 ] [dash pattern={on 4.5pt off 4.5pt}]  (357,97) .. controls (262.92,110.72) and (272.56,148.45) .. (291.81,173.48) ;
\draw [shift={(293,175)}, rotate = 231.34] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ,draw opacity=1 ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw    (425.5,95) .. controls (380.42,119.5) and (406.41,202.58) .. (422.53,224.72) ;
\draw [shift={(423.5,226)}, rotate = 231.34] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw    (497.5,95) .. controls (564.48,91.06) and (565,230.71) .. (567.87,269.31) ;
\draw [shift={(568,271)}, rotate = 265.24] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw    (39.5,359) .. controls (-12.71,369.84) and (17.07,323.43) .. (32.31,293.36) ;
\draw [shift={(33,292)}, rotate = 116.57] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw    (122,215) .. controls (128.9,254.4) and (131.91,269.54) .. (159.71,289.1) ;
\draw [shift={(161,290)}, rotate = 214.59] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw    (257,197) .. controls (258.95,236) and (277.06,270.25) .. (293.72,290.47) ;
\draw [shift={(295,292)}, rotate = 229.64] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw    (386,177) .. controls (392.86,226) and (408.36,270.2) .. (425.92,290.77) ;
\draw [shift={(427,292)}, rotate = 228.01] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw    (517,229) .. controls (524.76,253.25) and (540.05,269.03) .. (552.82,282.73) ;
\draw [shift={(554,284)}, rotate = 227.12] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw   (514,336) -- (584,336) -- (584,376) -- (514,376) -- cycle ;

@@

\draw   (95,338) -- (136.5,338) -- (136.5,378) -- (95,378) -- cycle ;

@@

\draw   (136.5,338) -- (265.5,338) -- (265.5,378) -- (136.5,378) -- cycle ;

@@

\draw   (265.5,338) -- (305,338) -- (305,378) -- (265.5,378) -- cycle ;
\draw   (558,284.5) .. controls (558,279.25) and (562.25,275) .. (567.5,275) .. controls (572.75,275) and (577,279.25) .. (577,284.5) .. controls (577,289.75) and (572.75,294) .. (567.5,294) .. controls (562.25,294) and (558,289.75) .. (558,284.5) -- cycle ; \draw   (558,284.5) -- (577,284.5) ; \draw   (567.5,275) -- (567.5,294) ;

@@

\draw    (568,298) .. controls (584.66,309.76) and (586.91,318.64) .. (565.35,335.93) ;
\draw [shift={(564,337)}, rotate = 321.95] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

% Text Node
\draw (11,14) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Virtual}};
% Text Node
\draw (570,395) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Physical}};
% Text Node
\draw (104,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 63}};
% Text Node
\draw (175,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 48}};
% Text Node
\draw (245,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 39}};
% Text Node
\draw (314,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 30}};
% Text Node
\draw (385,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 21}};
% Text Node
\draw (454,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 12}};
% Text Node
\draw (523,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 0}};
% Text Node
\draw (132,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Sign}};
% Text Node
\draw (195,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L4 Offset}};
% Text Node
\draw (264,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L3 Offset}};
% Text Node
\draw (336,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L2 Offset}};
% Text Node
\draw (406,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L1 Offset}};
% Text Node
\draw (472,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Page Offset}};
% Text Node
\draw (67,63) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize va}};
% Text Node
\draw (48,203) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L4 Entry}};
% Text Node
\draw (180,189) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L3 Entry}};
% Text Node
\draw (311,167) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L2 Entry}};
% Text Node
\draw (444,223) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L1 Entry}};
% Text Node
\draw (51,347) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize root}};
% Text Node
\draw (465,349) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize page}};
% Text Node
\draw (529,346) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 4K Data}};
% Text Node
\draw (133,386) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 51}};
% Text Node
\draw (258,386) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 12}};
% Text Node
\draw (163,350) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Base Addr}};

\end{tikzpicture}
          \caption{x86-64 page table lookups.}
        \label{fig:pagetables}
    \end{subfigure}
    \vfill
\begin{minipage}{.5\textwidth}
%    \begin{figure}
\begin{lstlisting}[mathescape,escapeinside={(*}{*)},basicstyle=\footnotesize]
pte_t *pte_get_next_table(pte_t *entry) {
  pte_t *next;
  /* Check if entry is populated, allocate if needed(* \label{line:check_entry_presentC} *)(* \label{line:mask_presentC} *)(*\label{line:after_concluding_qfrac1C}*)*/
  if (!entry->present(*\label{line:read_entry_contentsC}*) (*\label{line:conditional_childrenC}*)) {
    pte_initialize(entry); // Alloc empty table for next level (*\label{line:pass_addrof_nextC}*)(*\label{line:alloc_path_startC}*)(* \label{line:call_to_pte_initializeC} *)
    entry->present = 1; // Mark valid (*\label{line:install_new_entryC}*)
    /* Key invariant for child access now holds */
  } (*\label{line:end_of_allocation_pathCB}*)
  /*Physical to virtual address mapping code segment starts here (*\label{line:finalpieceSB}*)*/
  uintptr_t next_phys_addr = PTE_PFN_TO_ADDR(entry->pfn)
  uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr); (*\label{line:p2vCB}*);
  next = (pte_t *) next_virt_addr;(*\label{line:finalpieceEB}*)
  return next;
}
\end{lstlisting}
\end{minipage}
\hfill\hfill
\begin{minipage}{.4\textwidth}
\begin{lstlisting}[mathescape,escapeinside={(*}{*)}, basicstyle=\footnotesize]
pte_t *walkpgdir(pte_t *root, void *va){ 
  pte_t *l4_entry = &root[L4Offset(va)];
  pte_t *l3 = pte_next_table(l4_entry);
  pte_t *l3_entry = &l3[L3Offset(va)];
  pte_t *l2 = pte_next_table(l3_entry);
  pte_t *l2_entry = &l2[L2Offset(va)];  
  pte_t *l1 = pte_next_table(l2_entry);
  pte_t *l1_entry = &l1[L1Offset(va)];  
  return l1_entry
}
\end{lstlisting}
%          \caption{Walking page-table tree to locate L1 entry C-source code.}
%        \label{fig:enter-label}
 %   \end{figure}
\end{minipage}
%\end{framed}
    \vspace{-2.5em}
    \caption{A kernel implementation of x86-64 page table lookups.}
    \label{fig:pagetablescode}
    \vspace{-1em}
\end{figure}

@@

Of particular interest for reasoning about virtual memory are modalities that permit \emph{naming} the alternate 
circumstances, prominently \emph{hybrid} modal logics~\cite{blackburn1995hybrid,areces2001hybrid}, which come equipped 
with assertions of the form $[\ell](P)$ indicating that $P$ is true in the specific alternate circumstance (Kripke world)
 named by the \emph{nominal} $\ell$. Note that a distinctive property of hybrid logics is that, rather than hiding
the points at which a modal assertion is evaluated inside the modality's definition, the choice of what world a modalized
assertion should be true in is chosen \emph{in the assertion itself}. This allows assertions to talk about not simply whether some other assertion
is true in some possible future or past world related in a fixed way to the current world, but to talk about \emph{arbitrary}
other worlds.

>>>>

In typical system configurations, all memory addresses seen by programs running on modern computers are \emph{virtualized}: the address observed by a running
program generally will not correspond directly to the physical location in memory, and may not even correspond to a physical location that \emph{exists} 
in the machine. Instead, these \emph{virtual} addresses are translated to \emph{physical} addresses that correspond directly to locations in RAM. On most 
modern architectures, this translation is performed through cooperation of the hardware and OS kernel: while executing an instruction that dereferences a 
(virtual) address, the CPU's \emph{memory management unit} (\textsc{MMU}) hardware performs \emph{address translation},
resulting in a physical address used to access the cache\footnote{Technically, for performance reasons most caches are indexed with parts of the virtual 
address, but tagged with the physical data addresses, so cache lookups and address translations can proceed in parallel.} and/or memory-bus.
\setlength{\columnseprule}{4pt}
\begin{figure}\scriptsize
%    \begin{framed}
%\column{.59\textwidth}    
\begin{subfigure}\scriptsize

@@

\begin{tikzpicture}[x=0.75pt,y=0.75pt,yscale=-0.55,xscale=0.55]
%uncomment if require: \path (0,403); %set diagram left start at 0, and has height of 403

@@

\draw   (33,133) -- (123,133) -- (123,284) -- (33,284) -- cycle ;

@@

\draw   (166,133) -- (256,133) -- (256,284) -- (166,284) -- cycle ;

@@

\draw   (295,133) -- (385,133) -- (385,284) -- (295,284) -- cycle ;

@@

\draw   (427,133) -- (517,133) -- (517,284) -- (427,284) -- cycle ;

@@

\draw   (115,46) -- (185,46) -- (185,86) -- (115,86) -- cycle ;

@@

\draw   (185,46) -- (255,46) -- (255,86) -- (185,86) -- cycle ;

@@

\draw   (255,46) -- (325,46) -- (325,86) -- (255,86) -- cycle ;

@@

\draw   (325,46) -- (395,46) -- (395,86) -- (325,86) -- cycle ;

@@

\draw   (395,46) -- (465,46) -- (465,86) -- (395,86) -- cycle ;

@@

\draw   (464.5,46) -- (555.5,46) -- (555.5,86) -- (464.5,86) -- cycle ;

@@

\draw    (211.5,88) .. controls (-36.44,90.94) and (15.75,182.24) .. (30.17,205.65) ;
\draw [shift={(31,207)}, rotate = 238.24] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw   (33,183) -- (123,183) -- (123,223) -- (33,223) -- cycle ;

@@

\draw    (289.5,87) .. controls (115.15,102.76) and (137.77,167.03) .. (163.33,189.98) ;
\draw [shift={(164.5,191)}, rotate = 220.24] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw   (166,169) -- (256,169) -- (256,209) -- (166,209) -- cycle ;

@@

\draw  [dash pattern={on 4.5pt off 4.5pt}] (295,148) -- (385,148) -- (385,188) -- (295,188) -- cycle ;

@@

\draw   (427,204) -- (517,204) -- (517,244) -- (427,244) -- cycle ;

@@

\draw [color={rgb, 255:red, 0; green, 0; blue, 0 }  ,draw opacity=1 ] [dash pattern={on 4.5pt off 4.5pt}]  (357,89) .. controls (262.92,102.72) and (272.56,140.45) .. (291.81,165.48) ;
\draw [shift={(293,167)}, rotate = 231.34] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ,draw opacity=1 ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw    (425.5,87) .. controls (380.42,111.5) and (406.41,194.58) .. (422.53,216.72) ;
\draw [shift={(423.5,218)}, rotate = 231.34] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw    (497.5,87) .. controls (564.48,83.06) and (567.43,193.61) .. (570.37,231.33) ;
\draw [shift={(570.5,233)}, rotate = 265.24] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw    (36.5,330) .. controls (-15.71,340.84) and (16.98,314.8) .. (32.31,285.35) ;
\draw [shift={(33,284)}, rotate = 116.57] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw    (122,207) .. controls (128.9,246.4) and (131.91,261.54) .. (159.71,281.1) ;
\draw [shift={(161,282)}, rotate = 214.59] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw    (257,189) .. controls (258.95,228) and (277.06,262.25) .. (293.72,282.47) ;
\draw [shift={(295,284)}, rotate = 229.64] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw    (386,169) .. controls (392.86,218) and (408.36,262.2) .. (425.92,282.77) ;
\draw [shift={(427,284)}, rotate = 228.01] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw    (517,221) .. controls (535.13,236.68) and (527.81,247.56) .. (555.75,240.46) ;
\draw [shift={(557.5,240)}, rotate = 165.07] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

@@

\draw   (511,313) -- (581,313) -- (581,353) -- (511,353) -- cycle ;

@@

\draw   (95,313) -- (136.5,313) -- (136.5,353) -- (95,353) -- cycle ;

@@

\draw   (136.5,313) -- (265.5,313) -- (265.5,353) -- (136.5,353) -- cycle ;

@@

\draw   (265.5,313) -- (305,313) -- (305,353) -- (265.5,353) -- cycle ;
\draw   (560,243.5) .. controls (560,238.25) and (564.25,234) .. (569.5,234) .. controls (574.75,234) and (579,238.25) .. (579,243.5) .. controls (579,248.75) and (574.75,253) .. (569.5,253) .. controls (564.25,253) and (560,248.75) .. (560,243.5) -- cycle ; \draw   (560,243.5) -- (579,243.5) ; \draw   (569.5,234) -- (569.5,253) ;

@@

\draw    (569,258) .. controls (585.66,269.76) and (587.91,278.64) .. (566.35,295.93) ;
\draw [shift={(565,297)}, rotate = 321.95] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

% Text Node
\draw (21,20) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize Virtual}};
% Text Node
\draw (569,375) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize Physical}};
% Text Node
\draw (102,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 63}};
% Text Node
\draw (173,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 48}};
% Text Node
\draw (243,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 39}};
% Text Node
\draw (312,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 30}};
% Text Node
\draw (383,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 21}};
% Text Node
\draw (452,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 12}};
% Text Node
\draw (547,16) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 0}};
% Text Node
\draw (126,54) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize Sign}};
% Text Node
\draw (188,54) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L4 Offset}};
% Text Node
\draw (257,54) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L3 Offset}};
% Text Node
\draw (328,54) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L2 Offset}};
% Text Node
\draw (398,55) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L1 Offset}};
% Text Node
\draw (471,54) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize Page Offset}};
% Text Node
\draw (87,52) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize va}};
% Text Node
\draw (46,192) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L4 Entry}};
% Text Node
\draw (178,178) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L3 Entry}};
% Text Node
\draw (309,156) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L2 Entry}};
% Text Node
\draw (442,212) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L1 Entry}};
% Text Node
\draw (59,322) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize root}};
% Text Node
\draw (471,321) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize page}};
% Text Node
\draw (517,321) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 4K Data}};
% Text Node
\draw (131,358) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 51}};
% Text Node
\draw (256,358) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 12}};
% Text Node
\draw (161,322) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize Base Addr}};

\end{tikzpicture}

          \caption{x86-64 page table lookups.}
        \label{fig:pagetables}
    \end{subfigure}
    \vfill
\begin{minipage}{.5\textwidth}
%    \begin{figure}
\begin{lstlisting}[mathescape,escapeinside={(*}{*)},basicstyle=\scriptsize]
pte_t *pte_get_next_table(pte_t *entry) {
  pte_t *next;
  /* Check if entry is populated, allocate if needed(* \label{line:check_entry_presentC} *)(* \label{line:mask_presentC} *)(*\label{line:after_concluding_qfrac1C}*)*/
  if (!entry->present(*\label{line:read_entry_contentsC}*) (*\label{line:conditional_childrenC}*)) {
    pte_initialize(entry); // Alloc empty table for next level (*\label{line:pass_addrof_nextC}*)(*\label{line:alloc_path_startC}*)(* \label{line:call_to_pte_initializeC} *)
    entry->present = 1; // Mark valid (*\label{line:install_new_entryC}*)
    /* Key invariant for child access now holds */
  } (*\label{line:end_of_allocation_pathCB}*)
  /*Physical to virtual address mapping code segment starts here (*\label{line:finalpieceSB}*)*/
  uintptr_t next_phys_addr = PTE_PFN_TO_ADDR(entry->pfn)
  uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr); (*\label{line:p2vCB}*);
  next = (pte_t *) next_virt_addr;(*\label{line:finalpieceEB}*)
  return next;
}
\end{lstlisting}
\end{minipage}
\hfill\hfill
\begin{minipage}{.4\textwidth}
\begin{lstlisting}[mathescape,escapeinside={(*}{*)}, basicstyle=\scriptsize]
pte_t *walkpgdir(pte_t *root, void *va){ 
  pte_t *l4_entry = &root[L4Offset(va)];
  pte_t *l3 = pte_next_table(l4_entry);
  pte_t *l3_entry = &l3[L3Offset(va)];
  pte_t *l2 = pte_next_table(l3_entry);
  pte_t *l2_entry = &l2[L2Offset(va)];  
  pte_t *l1 = pte_next_table(l2_entry);
  pte_t *l1_entry = &l1[L1Offset(va)];  
  return l1_entry
}
\end{lstlisting}
%          \caption{Walking page-table tree to locate L1 entry C-source code.}
%        \label{fig:enter-label}
 %   \end{figure}
\end{minipage}
%\end{framed}
    \vspace{-2.5em}
    \caption{A kernel implementation of x86-64 page table lookups.}
    \label{fig:pagetablescode}
    \vspace{-1em}
\end{figure}

@@

Of particular interest for reasoning about virtual memory are modalities that permit \emph{naming} the alternate 
circumstances, prominently \emph{hybrid} modal logics~\cite{blackburn1995hybrid,areces2001hybrid}, which come equipped 
with assertions of the form $[\ell](P)$ indicating that $P$ is true in the specific alternate circumstance (Kripke world)
 named by the \emph{nominal} $\ell$. Note that a distinctive property of hybrid logics is that, rather than hiding
the points at which a modal assertion is evaluated inside the modality's definition, the choice of what world a modalized
assertion should be true in is chosen \emph{in the assertion itself}. This allows assertions to talk about not simply whether some other assertion
is true in some possible future or past world related in a fixed way to the current world, but to talk about \emph{arbitrary}
other worlds.
\looseness=-1
